---
title: "Nicole_Analysis"
author: "Nicole Ham"
date: "`r Sys.Date()`"
output: github_document
---

```{r}
library(tidyverse)
library(dplyr)
library(stringr)
```

```{r}
habits_df <- read_csv("student_habits_performance.csv")
social_df <- read_csv("smmh.csv")

habits_df <- habits_df %>%
  mutate(
    age = as.numeric(age),
    soc_range = cut(
      social_media_hours,
      breaks = c(-Inf, 1, 2, 3, 4, Inf),
      labels = c("0-1 hours", "1-2 hours", "2-3 hours", "3-4 hours", "4+ hours"),
      right = TRUE),
    age_range = cut(
      age,
      breaks = c(-Inf, 17, 21, 25, Inf),
      labels = c("<18", "18-21", "22-25", "26+"),
      right = TRUE))

social_df_clean <- social_df %>%
  mutate(
    age = as.numeric(`1. What is your age?`),
    soc_hrs = tolower(trimws(`8. What is the average time you spend on social media every day?`)),
    soc_num = as.numeric(str_extract(soc_hrs, "\\d+")),
    soc_range = cut(
      soc_num,
      breaks = c(-Inf, 1, 2, 3, 4, Inf),
      labels = c("0-1 hours", "1-2 hours", "2-3 hours", "3-4 hours", "4+ hours"),
      right = TRUE),
    age_range = cut(
      age,
      breaks = c(-Inf, 17, 21, 25, Inf),
      labels = c("<18", "18-21", "22-25", "26+"),
      right = TRUE))

merged_df <- inner_join(
  habits_df %>% select(-gender),
  social_df_clean,
  by = c("soc_range", "age_range"))

merged_df <- merged_df %>%
  select(-age.x, 
         -age.y, 
         -student_id, 
         -soc_hrs, 
         -soc_num, 
         -Timestamp, 
         -`1. What is your age?`,
         -`2. Gender`,
         -`3. Relationship Status`,
         -`5. What type of organizations are you affiliated with?`,
         -`6. Do you use social media?`,
         -`8. What is the average time you spend on social media every day?`,
         -parental_education_level)
```

```{r}
head(merged_df)
names(merged_df)
```

## Research Question: Is there a correlation between a student’s mental health and their exam score?
```{r}
ggplot(merged_df, aes(x = mental_health_rating, y = exam_score)) +
  geom_point(color = "steelblue",size = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_x_continuous(breaks = 1:10, limits = c(1, 10)) +
  labs(
    title = "Relationship Between Mental Health Rating and Exam Score",
    x = "Mental Health Rating",
    y = "Exam Score")
```

### Doing some experimenting here and creating a heatmap
```{r}
install.packages("corrplot")
```


```{r}
library(corrplot)

vars <- c("mental_health_rating", "sleep_hours", "exercise_frequency",
          "social_media_hours", "exam_score", "attendance_percentage", "study_hours_per_day")
subset_data <- merged_df[ , vars]
subset_data <- na.omit(subset_data)

head(subset_data)

cor_matrix <- cor(subset_data, method = "pearson")
corrplot(cor_matrix, 
         method = "color", 
         type = "lower",
         addCoef.col = "black", 
         tl.col = "black", 
         tl.srt = 45)
```
How to interpret the data:
Greater than 1 - positive correlation
0 - no relationship
Lower than 1 - negative correlation

i.e
Exam score and social media hours has a negative correlation of -0.16, meaning that more social media hours means a correlation of a lower exam score
Exam score and mental health has a positive correlation of 0.32, meaning that the higher the mental health score, the better students do. 

```{r}
vars <- c("sleep_hours", "exercise_frequency", "social_media_hours", 
          "study_hours_per_day", "attendance_percentage", "exam_score")

long_data <- merged_df %>%
  select(all_of(vars), mental_health_rating) %>%
  pivot_longer(cols = all_of(vars), names_to = "Variable", values_to = "Value")

# Step 2: Group by variable and value, then calculate mean mental health
avg_long <- long_data %>%
  group_by(Variable, Value) %>%
  summarise(avg_mental_health = mean(mental_health_rating, na.rm = TRUE), .groups = "drop")

# Step 3: Line plot faceted by variable
ggplot(avg_long, aes(x = Value, y = avg_mental_health)) +
  geom_line(color = "black", size = 0.5) +
  geom_point(color = "blue", size = 0.6) +
  facet_wrap(~ Variable, scales = "free_x") +
  labs(title = "How Mental Health Rating Varies by Different Factors",
       x = "Value of Variable",
       y = "Average Mental Health Rating")
```


```{r}
vars <- c("sleep_hours", "exercise_frequency", "social_media_hours", 
          "study_hours_per_day", "attendance_percentage", "exam_score")

long_data <- merged_df %>%
  select(all_of(vars), mental_health_rating) %>%
  pivot_longer(cols = all_of(vars), names_to = "Variable", values_to = "Value")

# Step 2: Plot with correlation/trend lines
ggplot(long_data, aes(x = Value, y = mental_health_rating)) +
  geom_point(alpha = 0.4, color = "darkblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  facet_wrap(~ Variable, scales = "free_x") +
  labs(title = "Mental Health Rating vs. Various Factors",
       x = "Value of Variable",
       y = "Mental Health Rating")
```

